---
title: "DAT301 Lab 1"
name: "Madison(Dayeon) Kim"
output: pdf_document
date: "2026-08-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read the file

```{r}
df = read.csv("Absenteeism_at_work.csv",sep=";",header=TRUE)
```

\vspace{0.5cm}

# 1. Plot the scatter plot of height vs. weight.

```{r}
x = df$Weight
y = df$Height

plot(x,y,pch=16, xlab= 'Weight (lbs)',
                 ylab = 'Height (cm)',
                 main = 'Height vs Weight')
```

There is no significant linear relationship between weight and height in absenteeism, thus showing that there's no correlation in height and weight.

\vspace{1cm}

# 2. Plot the histogram of hours of absences.

```{r}
hist(df$Absenteeism.time.in.hours, breaks=12, col='lightgreen',
     xlab='Hours of Absences',
     ylab='Frequency',
     main="Histogram of Absenteeism Hours")

```

Hours of absences are right-skewed, meaning majority of absences are occurring at lower hours of absences.

\vspace{1cm}

# 3. Plot the histogram of age of a person corresponding to each absence.

```{r}
hist(df$Age, breaks=12, col='lightblue',
     xlab='Age',
     ylab='Frequences',
     main = 'Histogram of Ages')
```

The histogram does not show a clear pattern, but it is slightly right-skewed, because there are fewer observations at higher ages.

\vspace{1cm}

# 4. Plot the bar plot of hours by month.

```{r}
table(df$Month.of.absence)
```

```{r}
monthly_hours = tapply(df$Absenteeism.time.in.hours, df$Month.of.absence, sum, na.rm=T)

barplot(monthly_hours,
        col='brown',
        main = "Total Absence Hours by Month",
        xlab = "Month",
        ylab = "Total Absence Hours")
```

Three absences were recorded without a specified month. The highest number of absenteeism hours was recorded in March, followed by July, while the lowest was recorded in January.

\vspace{1cm}

# 5. Plot the box plots of hours by social smoker variable.

```{r}
table(df$Social.smoker)
unique(df$Absenteeism.time.in.hours)
```

```{r}

boxplot(df$Absenteeism.time.in.hours ~ df$Social.smoker,
        data = df, 
        xlab= "Absenteeism Hours by Social Smoker Status",
        ylab= "Absent Hours",
        names = c("Non-Smoker","Smoker"),
        col = c("lightblue","pink"),
        main = "Box Plot of Hours by Smoking Status")

legend(
  "topright", 
  legend = c("Non-Smoker", "Smoker"), 
  fill = c("lightblue", "pink")
)

```

More outliers are recorded in the Non-Smoker group, reaching up to more than 110 hours.

\vspace{1cm}

# 6. Plot the box plots of hours by social drinker variable.

```{r}

boxplot(df$Absenteeism.time.in.hours ~ df$Social.smoker,
        data = df, 
        horizontal=T,
        xlab= "Absenteeism Hours by Social Smoker Status",
        ylab= "Absent Hours",
        ylim=c(0,10),
        names = c("Non-Smoker","Smoker"),
        col = c("turquoise","orange"),
        main = "Box Plot of Hours by Smoking Status")

abline(v=median(df$Absenteeism.time.in.hours, na.rm=T), col='red',lty='dashed', lwd=2)

legend(
  "topright",
  legend = c("Non-Smoker","Smoker"),
  fill = c("turquoise","orange"))

```

While the Non-Smoker group has a slightly lower median absenteeism hours than the Smoker group, both groups appear to have identical 25th and 75th percentiles.

The Non-Smoker median falls right at the entire median and Smoker median is slightly away from the entire median.
